Bayesian Learning of Degenerate Linear~Gaussian~State~Space~Models using~Markov~Chain~Monte~Carlo

### 2016-01-15 notes/reviews/response_to_reviews.tex

\begin{document}

\maketitle

Thank you for taking the time to review our paper. Here is our responses to the points raised.

\meta{
\begin{itemize}
 \item Add trace plots for the transition matrix.
 \item change the observation noise so the rank is not so certain.
 \item Add rank posterior histograms.
 \item leave out parameter conditional moves?
 \item plots
 \item Reviewer 3
 \item Proof read, aiming to (i) clarify (2) correct typos (c) define everything when first used
\end{itemize}

@@

\begin{review}
3.  The $H$ and $R$ matrices are assumed to be known in the manuscript, which may not be the case in practice. It is suggested to consider the estimation of $H$ and $R$ in the degenerate model.
\end{review}

\begin{response}
 See response to the similar comment by reviewer 1.
\end{response}

\begin{review}
4.  This manuscript uses Givens decomposition to factorize matrix $Q$, and in section III B, it states that matrix $U$ in Givens decomposition is uniquely determined. It should be explained that how to determine matrix $U$ and the necessity of Givens decomposition.
\end{review}

\begin{response}
 We have added more explanation of the Givens decomposition, and also algorithm pseudo-code for its calculation.
\end{response}

### 2016-01-15 notes/paper-degenerate-linear-systems.tex
Added Givens decomposition pseudo-code and additional explanation

A further factorisation will also be useful, which we will refer to as the Givens decomposition (on account of its method of computation),
%
\begin{IEEEeqnarray}{rCl}
 \lgtv &=& \tvrot \tvfull \tvrot\tr     , \label{eq:givens_decomposition} %&=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix}
\end{IEEEeqnarray}
%
where $\tvfull$ is an $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors subject to certain constraints. The transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective. See appendix~\ref{app:givens-factorisation} for details.

@@

The hyperparameters were set to generally uninformative values. For the full rank model,
%
\begin{IEEEeqnarray}{rClCrCl}
 \priordof &=& \lsd - 1 & \qquad & \priorscalematrix &=& \priordof \idmat \nonumber \\
 \priormeanmatrix &=& \zmat & \qquad & \priorcolumnvariance &=& 100 \idmat      .
\end{IEEEeqnarray}
%
Similarly, for the degenerate model,
%
\begin{IEEEeqnarray}{rClCrCl}
 \priorscalematrixbase &=& \idmat & \qquad & \priormeanmatrix &=& \zmat \nonumber \\
 \priorcolumnvariance &=& 100 \idmat & \qquad & \priortypval &=& 1 \nonumber \\
 \priorrankweight{i} &=& \frac{1}{\lsd}     .
\end{IEEEeqnarray}
%
The move probabilities were set to,
%
\begin{IEEEeqnarray}{rCl}
 P(\rk \to\rk+1) = P(\rk\to\rk+1) = \half     .
\end{IEEEeqnarray}
\meta{CHECK THIS!}

@@

A Givens rotation matrix has the following structure,
%
\begin{IEEEeqnarray}{rCl}
 \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
                                                    \sin(\givrot{}) & k=i,l=j \\
                                                    -\sin(\givrot{}) & k=j,l=i \\
                                                    0 & \text{ otherwise}     ,
                                                 \end{cases}
\end{IEEEeqnarray}
%
where $\givrot{} \in [-\pi/2,\pi/2]$ is a rotation in the plane of the $i$ and $j$ coordinate directions. Any orthogonal $d\times d$ matrix may be written as a product of $\half d(d-1)$ such rotations in the following way \cite{Anderson1987},
%
\begin{IEEEeqnarray}{rCl}
\tvvec = \tvsign &\times& \left[ \givmat{1}{2}{\givrot{1,2}}\right] \times \dotsm \nonumber\\
&\times& \left[\givmat{1}{d-1}{\givrot{1,d-1}} \dotsm \givmat{d-2}{d-1}{\givrot{d-2,d-1}} \right] \nonumber\\
&\times& \left[\givmat{1}{d}{\givrot{1,d}} \dotsm \givmat{d-1}{d}{\givrot{d-1,d}} \right] \label{eq:standard_givens}     ,
\end{IEEEeqnarray}
%
where $\tvsign$ is a diagonal matrix in which the diagonal elements are $\pm1$. This factorisation may be derived (and also calculated) by repeatedly left-multiplying the original matrix by a (transposed) Givens rotation matrix such that one element is set to $0$. The matrix remaining once all the elements below the diagonal have been eliminated is $\tvsign$, and the factorisation above results from a straightforward rearrangement. See \cite{Anderson1987} for details. This factorisation is not unique. Any order of Givens matrices may be used provided each operation does not interfere with the elements zeroed by the preceding operations. Furthermore, we can also zero an element of the matrix by right multiplication by a Givens matrix, so the rotations may be split into two blocks, one each side of the sign matrix.

@@

\begin{IEEEeqnarray}{rCl}
 \tvvec & = & \tvcross \tvsign \tvrow      ,
\end{IEEEeqnarray}
%
where
%
\begin{IEEEeqnarray}{rCl}
 \tvrow  & = & \left[\givmat{1}{2}{\givrot{1,2}}\right] \times \dotsm \nonumber \\
 & & \qquad\qquad \times \left[\givmat{1}{r}{\givrot{1,r}} \dots \givmat{r-1}{r}{\givrot{r-1,r}}\right] \\
 \tvcross & = & \left[\givmat{1}{r+1}{\givrot{1,r+1}}\dots\givmat{1}{d}{\givrot{1,d}}\right] \times \dotsm \nonumber \\
 & & \qquad\qquad \times \left[\givmat{r}{r+1}{\givrot{r,r+1}} \dots \givmat{r}{d}{\givrot{r,d}}\right]      ,
\end{IEEEeqnarray}
%
and $\tvsign$ is as before, with the following structure,
%
\begin{IEEEeqnarray}{rCl}
 \tvsign & = & \begin{bmatrix}
                  \tvsigndiag \\
                  \zmat[(d-r)\times r]
                 \end{bmatrix} \nonumber      .
\end{IEEEeqnarray}

Now consider a rank-$r$ positive semi-definite matrix $\psd$ for which the non-singular part of the eigendecomposition is,
%
\begin{IEEEeqnarray}{rCl}
 \psd &=& \tvvec \tvval \tvvec\tr     . 
\end{IEEEeqnarray}
%
Factorising the matrix of eigenvectors using the Givens method, we find,
%
\begin{IEEEeqnarray}{rCl}
 \psd & = & \tvcross \tvsign \tvrow \tvval \tvrow\tr \tvsign\tr \tvcross\tr  \nonumber \\
 & = & \tvcross \begin{bmatrix}
                  \tvsigndiag \tvrow \tvval \tvrow\tr \tvsigndiag\tr & \zmat \\
                  \zmat & \zmat
                \end{bmatrix} \tvcross\tr \nonumber \\
 & = & \tvrot \tvfull \tvrot\tr \nonumber      ,
\end{IEEEeqnarray}
%
in which $\tvrot$ consists of the first $r$ columns of $\tvcross$, and,
%
\begin{IEEEeqnarray}{rCl}
 \tvfull &=& \tvsigndiag \tvrow \tvval \tvrow\tr \tvsigndiag\tr    .
\end{IEEEeqnarray}
%
Comparing with \eqref{eq:standard_givens}, we see that $\tvsigndiag \tvrow$ is an unconstrained $r\times r$ orthogonal matrix and hence $\tvfull$ is a unique positive definite matrix.

>>>>

A second decomposition will also be useful. Any positive semi-definite can be factorised in the following way,
%
\begin{IEEEeqnarray}{rCl}
 \lgtv &=& \tvrot \tvfull \tvrot\tr     , \label{eq:givens_decomposition}
\end{IEEEeqnarray}
%
where $\tvfull$ is an $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors with a particular structure. The transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective. This factorisation is constructed using Givens rotations and thus we refer to it as the Givens decomposition. A Givens rotation matrix has the following structure,
%
\begin{IEEEeqnarray}{rCl}
 \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
                                                    \sin(\givrot{}) & k=i,l=j \\
                                                    -\sin(\givrot{}) & k=j,l=i \\
                                                    0 & \text{ otherwise}     ,
                                                 \end{cases}
\end{IEEEeqnarray}
%
where $\givrot{} \in [-\pi/2,\pi/2]$ is a rotation in the plane of the $i$ and $j$ coordinate directions.

The Givens decomposition is important because $\tvrot$ is uniquely determined by the state trajectory, whereas $\tvfull$ is free to vary. We provide here the algorithm for conducting the Givens decomposition. The computational complexity is $\bigo{\rk\times(\lsd-\rk)}$. (Note that multiplication by a Givens matrix is $\bigo{1}$ due to sparsity.) Details and further discussion is included in appendix~\ref{app:givens-factorisation}.

\begin{algorithm}
\onehalfspacing
\small
\begin{algorithmic}[1]
 \REQUIRE Eigvenvalues and eigenvectors $\tvval,\tvvec$ of positive semi-definite matrix $\lgtv$.
 \STATE Set $\tvcross \leftarrow \idmat[\lsd\times\lsd]$, $\tvrow \leftarrow \idmat[\rk\times\rk]$, $\tvsign \leftarrow \tvvec$.
 \FOR{$i = \rk,\dots,1$}
   \FOR{$j = 1,\dots,\rk-1$}
      \STATE Calculate the rotation $\givrot{i,j} = \tan^{-1}(\tvsign_{i,j} / \tvsign_{i,i})$.
      \STATE Construct the Givens matrix $\givmat{i}{j}{\givrot{i,j}}$.
      \STATE Set $\tvsign \leftarrow \tvsign \: \givmat{i}{j}{\givrot{i,j}}$, $\tvrow \leftarrow \givmat{i}{j}{\givrot{i,j}}\tr \tvrow$.
   \ENDFOR
 \ENDFOR
  \FOR{$j = 1,\dots,\rk$}
   \FOR{$i = \lsd,\dots,\rk+1$}
      \STATE Calculate the rotation $\givrot{i,j} = \tan^{-1}(\tvsign_{j,j} / \tvsign_{i,j})$.
      \STATE Construct the Givens matrix $\givmat{i}{j}{\givrot{i,j}}$.
      \STATE Set $\tvsign \leftarrow \givmat{i}{j}{\givrot{i,j}} \tvsign$, $\tvcross \leftarrow \tvcross \givmat{i}{j}{\givrot{i,j}}\tr$.
   \ENDFOR
 \ENDFOR
 \STATE Extract $\tvsigndiag$, the first $\rk$ rows of $\tvsign$, and $\tvrot$, the first $\rk$ columns of $\tvcross$.
 \STATE Calculate $\tvfull = \tvsigndiag \tvrow \tvval \tvrow\tr \tvsigndiag\tr$.
 \ENSURE $\tvrot, \tvfull$.
\end{algorithmic}
\caption{Givens decomposition}
\label{alg:givens-decomposition}
\end{algorithm}

@@

The hyperparameters were set to generally uninformative values. For the full rank model,
%
\begin{IEEEeqnarray}{rClCrCl}
 \priordof &=& \lsd - 1 & \qquad & \priorscalematrix &=& \priordof \idmat \nonumber \\
 \priormeanmatrix &=& \zmat & \qquad & \priorcolumnvariance &=& 100 \idmat      .
\end{IEEEeqnarray}
%
Similarly, for the degenerate model,
%
\begin{IEEEeqnarray}{rClCrCl}
 \priorscalematrixbase &=& \idmat & \qquad & \priormeanmatrix &=& \zmat \nonumber \\
 \priorcolumnvariance &=& 100 \idmat & \qquad & \priortypval &=& 1 \nonumber \\
 \priorrankweight{i} &=& \frac{1}{\lsd}     .
\end{IEEEeqnarray}
%
The move probabilities were set to,
%
\begin{IEEEeqnarray}{rCl}
 P(\rk \to\rk+1) = P(\rk\to\rk+1) = \half     .
\end{IEEEeqnarray}

@@

A Givens rotation matrix has the following structure,
%
\begin{IEEEeqnarray}{rCl}
 \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
                                                    \sin(\givrot{}) & k=i,l=j \\
                                                    -\sin(\givrot{}) & k=j,l=i \\
                                                    0 & \text{ otherwise}     ,
                                                 \end{cases}
\end{IEEEeqnarray}
%
where $\givrot{} \in [-\pi/2,\pi/2]$ is a rotation in the plane of the $i$ and $j$ coordinate directions. Any orthogonal $d\times d$ matrix may be written as a product of $\half d(d-1)$ such rotations in the following way \cite{Anderson1987},
%
\begin{IEEEeqnarray}{rCl}
\tvvec = \tvsigndiag &\times& \left[ \givmat{1}{2}{\givrot{1,2}}\right] \times \dotsm \nonumber\\
&\times& \left[\givmat{1}{d-1}{\givrot{1,d-1}} \dotsm \givmat{d-2}{d-1}{\givrot{d-2,d-1}} \right] \nonumber\\
&\times& \left[\givmat{1}{d}{\givrot{1,d}} \dotsm \givmat{d-1}{d}{\givrot{d-1,d}} \right] \label{eq:standard_givens}     ,
\end{IEEEeqnarray}
%
where $\tvsigndiag$ is a diagonal matrix in which the diagonal elements are $\pm1$. This factorisation may be derived (and also calculated) by repeatedly left-multiplying the original matrix by a (transposed) Givens rotation matrix such that one element is set to $0$. The matrix remaining once all the elements below the diagonal have been eliminated is $\tvsigndiag$, and the factorisation above results from a straightforward rearrangement. See \cite{Anderson1987} for details. This factorisation is not unique. Any order of Givens matrices may be used provided each operation does not interfere with the elements zeroed by the preceding operations. Furthermore, we can also zero an element of the matrix by right multiplication by a Givens matrix, so the rotations may be split into two blocks, one each side of the sign matrix.

@@

\begin{IEEEeqnarray}{rCl}
 \tvvec & = & \tvcross \tvsign \tvrow      ,
\end{IEEEeqnarray}
%
where
%
\begin{IEEEeqnarray}{rCl}
 \tvrow  & = & \left[\givmat{1}{2}{\givrot{1,2}}\right] \times \dotsm \nonumber \\
 & & \qquad\qquad \times \left[\givmat{1}{r}{\givrot{1,r}} \dots \givmat{r-1}{r}{\givrot{r-1,r}}\right] \\
 \tvcross & = & \left[\givmat{1}{r+1}{\givrot{1,r+1}}\dots\givmat{1}{d}{\givrot{1,d}}\right] \times \dotsm \nonumber \\
 & & \qquad\qquad \times \left[\givmat{r}{r+1}{\givrot{r,r+1}} \dots \givmat{r}{d}{\givrot{r,d}}\right]      ,
\end{IEEEeqnarray}
%
and $\tvsign$ has the following structure,
%
\begin{IEEEeqnarray}{rCl}
 \tvsign & = & \begin{bmatrix}
                  \tvsigndiag \\
                  \zmat[(d-r)\times r]
                 \end{bmatrix} \nonumber      .
\end{IEEEeqnarray}
%
where $\tvsigndiag$, as before, is diagonal with elements $\pm1$.

Finally consider a rank-$r$ positive semi-definite matrix $\psd$ for which the non-singular part of the eigendecomposition is,
%
\begin{IEEEeqnarray}{rCl}
 \psd &=& \tvvec \tvval \tvvec\tr     . 
\end{IEEEeqnarray}
%
Factorising the matrix of eigenvectors using the Givens method, we find,
%
\begin{IEEEeqnarray}{rCl}
 \psd & = & \tvcross \tvsign \tvrow \tvval \tvrow\tr \tvsign\tr \tvcross\tr  \nonumber \\
 & = & \tvcross \begin{bmatrix}
                  \tvsigndiag \tvrow \tvval \tvrow\tr \tvsigndiag\tr & \zmat \\
                  \zmat & \zmat
                \end{bmatrix} \tvcross\tr \nonumber \\
 & = & \tvrot \tvfull \tvrot\tr \nonumber      ,
\end{IEEEeqnarray}
%
in which $\tvrot$ consists of the first $r$ columns of $\tvcross$, and,
%
\begin{IEEEeqnarray}{rCl}
 \tvfull &=& \tvsigndiag \tvrow \tvval \tvrow\tr \tvsigndiag\tr    .
\end{IEEEeqnarray}
%
Comparing with \eqref{eq:standard_givens}, we see that $\tvsigndiag \tvrow$ can represent any $r\times r$ orthogonal matrix and hence $\tvfull$ is a unique positive definite matrix.
